---
title: "Homework 5"
author: "Christopher Tsz Hin Choi"
date: "2025-11-30"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(
  echo = TRUE,  
  results = "markup", 
  warning = FALSE,   
  message = FALSE,
  error = FALSE    
)
```


```{r}
library(ggplot2)
```

##### Import data

```{r}
homicides = read.csv('data/homicide-data.csv') |>
  dplyr::mutate(city_name = paste0(city, ", ",state))
```

Find the city with the highest homicide and convert to sf
```{r}
homicide.city.count = homicides |>
  dplyr::group_by(city, state) |>
  dplyr::summarise(count = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::arrange(-count)

head(homicide.city.count)
```

```{r}
chicago.homicides = homicides |>
  dplyr::filter(city == 'Chicago',
                state == 'IL') |>
  sf::st_as_sf(coords = c("lon", "lat"),
               crs = 4269) ## crs same as tigris object

chicago.homicides
```

Import chicago vector
```{r}

chicago.boundary = tigris::places(state = "IL", cb = T, class = "sf") |>
  dplyr::filter(NAME == "Chicago")

ggplot(data = chicago.boundary) +
  geom_sf()
```

Plot locations of homicides into chicago
```{r}
ggplot() +
  geom_sf(data = chicago.boundary) +
  geom_sf(data = chicago.homicides, colour = "red", alpha=0.3)
```

Trun map into heat map to show density
```{r}

chicago.homicides.coords <- chicago.homicides |>
  sf::st_coordinates() |>
  as.data.frame() |>
  dplyr::bind_cols(chicago.homicides)


ggplot() +
  ## Density heat map
  geom_density2d_filled(
    data = chicago.homicides.coords,
    aes(x = X, y = Y, fill = after_stat(level)),
    alpha = 0.85,
    contour_var = "ndensity"
  ) +

  scale_fill_viridis_d(
    option = "magma",
    direction = -1,
    name = "Density Level"
  ) +
  ## Chicago BOundary
  geom_sf(data = chicago.boundary, color = "orange", size = 1, alpha = 0) +

  labs(
    title = "Homicide Density Across Chicago",
    subtitle = "Kernel density estimate of homicide locations",
    caption = paste0("Total number of homicides: ", nrow(chicago.homicides))
  ) +

  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 13),
    plot.caption = element_text(size = 9, color = "grey40")
  )


ggsave('Homicide_Density_Accross_Chicago.png')

```


#### Interpretation:

* Areas with highest homicide density in the area are in the darkest color contour

* Each lighter contour shows areas with proportionally lower relative density to the darkest contour



### SOlved/unsovled locations

```{r}
ggplot(data = chicago.homicides) +
  geom_sf(aes(colour = victim_race), alpha = 0.3)
  
```